---
title: "Optim - in class example"
author: "Naim Rashid"
date: "2/11/2019"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
header_includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsthm}
include-before:
- '\newcommand{\bfm}[1]{\ensuremath{\mathbf{#1}}}'
- '\newcommand{\bdm}[1]{\ensuremath{\boldsymbol{#1}}}'
- '$\def \d \bfm{d}$'
- '$\def \e \bfm{e}$'
- '$\def \g \bfm{g}$'
- '$\def \I \bfm{I}$'
- '$\def \l \bfm{l}$'
- '$\def \M \bfm{M}$'
- '$\def \W \bfm{W}$'
- '$\def \y \bfm{y}$'
- '$\def \Y \bfm{Y}$'
- '$\def \z \bfm{z}$'
- '$\def \Omegab \bdm{\Omega}$'
- '$\def \pib \bdm{\pi}$'
- '$\def \thetab \bdm{\theta}$'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A statistician wanted to fit a model to some data she had collected recently.  Here she assumed that the number of fish caught by fisherman in a study were related to their age, the size of their boat in feet, and the size of their storage cooler in liters.  

She tried fitting a regular possion GLM but the fit did not seem right.  Instead, she wanted to fit a mixture model assuming that there are actually two latent subpopulations in the data. There she can model the mean of each subpopulation with separate poisson regression models. 

Lets write the general form of a mixture model with $K$ mixture components, where each component corresponds to a particular assumed subpopulation in the data.  We can then write log-likelihood for this mixture as the following,  assuming $K = 2$ and $p = 4$, so

$$l(\boldsymbol{\beta}_1, \boldsymbol{\beta}_2, \pi_1, \pi_2) = \sum_{i=1}^n \log\left( \sum_{k =1}^2 \pi_kf_k(y_i| \boldsymbol{X}_i, \boldsymbol{\beta}_k)\right),$$ 

where $f_k(y_i|\boldsymbol{X}_i, \boldsymbol{\beta}_k) = \frac{e^{-\lambda_{ik}}\lambda_{ik}^{y_i}}{y_i!}$, $\lambda_{ik} = \exp(\boldsymbol{X}_{i}\boldsymbol{\beta}_k)$, and $\sum_{k = 1}^K \pi_k = 1$.  In this example, $\boldsymbol{X}_i$ is a $1 \times 4$ vector pertaining to the intercept, age, log boat length in feet, and log cooler size in liters, and $\boldsymbol{\beta}_k = (\beta_0, \beta_1, \beta_2, \beta_3)^T$ are the regression coefficients component $k$ pertaining to $\boldsymbol{X}_i$.  

Find a way to maximize this model to obtain estimates for $\hat{\thetab} = (\hat{\boldsymbol{\beta}_1},\hat{\boldsymbol{\beta}_2})^T$ using the data from fish.txt.  

To simplify the problem, we assume that $\pi_1 = 0.533$ is fixed and known ($\pi_2 = 1-\pi_1$), and use the starting values and code below. 

```{r}
# read in data
fish = read.table("/Users/naimrashid/Dropbox/UNC_bios_line/Bios735/statcomp_src/em/fish.txt", header = T, sep = "\t")

# create common covariate matrix
X = model.matrix(~1 + age + boat_length + cooler, data = fish)

# scaling the predictors to simplify the maximization
X[,2:4] = scale(X[,2:4])

# starting values and parameters
tol = 10^-4
beta = c(
      2.324, 0,0,0,    # first component starting betas
      3.375, 0,0,0   # second component starting betas
      )
maxit =5000
pi1 = 0.533
pi2 = 1-pi
```





